The Development of a Collaborative Tool to Teach Debugging

### 2020-12-05 paper/paper.tex
Final Draft Finished

@@

\author{Sammy Furr}
\title{The Development of a Collaborative Tool to Teach Debugging}
\date{\today}

@@

\begin{titlepage}
  \maketitle
\end{titlepage}

\begin{abstract}
  TODO: write an abstract
\end{abstract}

@@

Students using all three examples would benefit greatly from
visualization tools to help understand pointer locations, the process
space, and data structures.  The steps necessary to implement
visualization tools in the collaborative debugger are discussed
further in (\ref{next}).
\par

@@

Before diving into the construction of the frontend webapp, it's
helpful to look at the steps two students would take to debug the
\lstinline{memoization} example program remotely.

@@

  \includegraphics[width=\textwidth]{memoization}
  \centering
  \caption{Debugging the Memoization Example Session}
  \label{debugger:overview}
\end{figure}

@@

Notice that the debugging window in the example shown above is called
``rr''.  rr is an enhancment to gdb that is foundational in the
collaborative debugger's ability to deterministically record and
replay program execution. rr and its role in the collaborative
debugger will be covered extensively in (\ref{rrpods}).  Users
interact with rr through the collaborative debugger's frontend.

@@

The frontend of the collaborative debugger must handle constant
changes in state as users participate in the process outlined above.
The frontend uses the following tools to manage state and display
output.

@@

@@

\subsection{RR Message Server}

@@

@@

@@

TODO

>>>>

% From https://tex.stackexchange.com/questions/89574/language-option-supported-in-listings

\definecolor{lightgray}{rgb}{.9,.9,.9}
\definecolor{darkgray}{rgb}{.4,.4,.4}
\definecolor{purple}{rgb}{0.65, 0.12, 0.82}

\lstdefinelanguage{JavaScript}{
  keywords={typeof, new, true, false, catch, function, return, null, catch, switch, var, if, in, while, do, else, case, break},
  keywordstyle=\color{blue}\bfseries,
  ndkeywords={class, export, boolean, throw, implements, import, this},
  ndkeywordstyle=\color{darkgray}\bfseries,
  identifierstyle=\color{black},
  sensitive=false,
  comment=[l]{//},
  morecomment=[s]{/*}{*/},
  commentstyle=\color{purple}\ttfamily,
  stringstyle=\color{red}\ttfamily,
  morestring=[b]',
  morestring=[b]"
}

@@

\author{A Senior Project submitted to\\The Division of Science, Mathematics, and Computing\\of\\Bard College\\\\by\\Sammy Furr}
\title{The Development of a Collaborative Tool to Teach Debugging}
\date{\today}

@@

\begin{titlepage}
  \linespread{1}
    \begin{center}

        \Huge
        The Development of a Collaborative Tool to Teach Debugging

        \vspace{4cm}

        \large
        A Senior Project submitted to\\
        The Division of Science, Mathematics, and Computing\\
        of\\
        Bard College\\
        \vspace{1cm}
        by\\
        Sammy Furr
        \vfill

        Annandale-on-Hudson, New York\\
        December, 2020
    \end{center}
\end{titlepage}

\begin{abstract}
  \linespread{1}
  Debugging is rarely formally taught, despite being used by
  programmers every day.  Research indicates that it is valuable to
  teach debugging, and suggests that teaching it collaboratively may
  be maximally effective.  The goal of this project is to create a
  collaborative debugger.  The debugger aims to be the ideal platform
  to teach and learn debugging.  This paper briefly reviews relevant
  literature which covers teaching debugging and that which focuses on
  teaching programming collaboratively.  Most of the paper is devoted
  to the design of the collaborative debugger.
\end{abstract}

\section{Acknowledgments}

\begin{itemize}
\item Keith O'Hara, my senior project advisor, for his constant
  thoughtfulness about my project, his advice about the future, and
  for giving me the oppurtunity to tutor for his classes.
\item Mar Hicks and Kyle Hale, professors whose classes at IIT changed
  my perspective on computer science.
\item All my professors at Bard, particuarly Olga Voronina, Wakako
  Suzuki, and Nate Shockey, whose encouragement of my interests
  outside of computer science has been instrumental in continuing my
  passion for it.
\item My family and friends, for their constant love and support.
\end{itemize}

@@

Students using all three examples would benefit greatly from
visualization tools to help understand pointer locations, the process
space, and data structures.  As discussed in \ref{vistools}, adding
vizualization tools to the collaborative debugger should be relatively
easy.
\par

@@

Before diving into the construction of the frontend webapp, it may be
helpful to look at the steps two students would take to debug the
\lstinline{memoization} example program remotely:

@@

  \includegraphics[width=\textwidth]{memoization}
  \centering
  \caption{Debugging the Memoization Example Session}
  \label{frontend:rrterm}
\end{figure}

@@

Notice that the debugging window in the example shown above is called
``rr''.  rr is an enhancment to gdb that is foundational in the
collaborative debugger's ability to deterministically record and
replay program execution. rr and its role in the collaborative
debugger will be covered extensively in (\ref{rrpods}).  Users
interact with rr through the collaborative debugger's frontend.  Care
has been taken to make the frontend as representative of rr as
possible so that users can learn to use rr/gdb by using the
collaborative debugger.

@@

The frontend of the collaborative debugger must handle constant
changes in state as users participate in the process outlined above.
The frontend uses the following tools to manage state and display
output:

@@

In order to easily integrate Monaco and Xterm.js into the React
frontend, React wrapper libraries were used.  These are
react-xterm\cite{reactxterm} and
react-monaco-editor\cite{reactmonaco}.

React makes it easy to create simple webapps with no build
configuration necessary through the use of the
\lstinline{create-react-app} command.  This creates a structure and
toolchain for a basic one-page application---suitable for the
collaborative debugger.  Once this structure was created and any
superfluous files/code were removed, it's wasn't necessary for the
purposes of the collaborative debugger to directly edit any files
except two: \lstinline{App.js} and \lstinline{App.css}.  The
pre-defined build process takes care of transforming the React code
defined in \lstinline{App.js} and the styles defined in
\lstinline{App.css} into portable Javascript, HTML, and CSS.
\par

\lstinline{App.js} consists of a primary React Component class,
\lstinline{Debugger}, as well as many secondary classes and functions.
More complex components of the frontend, such as the Xterm.js
interface to rr, are created as classes, while simple components like
buttons are defined as functions.  All classes contain a
\lstinline{render()} method responsible for returning HTML elements to
display based on the state contained in the class.
\lstinline{Debugger}'s \lstinline{render()} method is responsible for
displaying all other components, based on whether the user is logged
in, has selected a new session, etc.
\par

Many components ``lift state up'' to \lstinline{Debugger}.  This is
accomplished by passing methods from \lstinline{Debugger} in as
``props'' when elements are created in \lstinline{Debugger}'s
\lstinline{render()} method:

\begin{lstlisting}[language=Javascript,basicstyle=\linespread{0.5}\ttfamily,caption={Lifting State Up},captionpos=b]
  onLogin(name) {
    this.setState({user: name})
    .
    .
    .
  }
  // In Debugger.render().  This is JSX.
  login = <LoginForm onLogin={this.onLogin}/>;

  // In LoginForm.
  this.props.onLogin(data.name);
\end{lstlisting}

\lstinline{Debugger}'s \lstinline{onLogin(name)} method changes the
state of debugger to indicate that the user is logged in.  When an
instance of \lstinline{LoginForm} is created in
\lstinline{Debugger.render()}, \lstinline{onLogin(name)} is passed to
it as a ``prop''.  The code for logging in the user can be contained
in \lstinline{LoginForm}, but when it becomes necessary to change the
state of the whole application to indicate that the user is logged in,
\lstinline{LoginForm} calls \lstinline{Debugger.onLogin(name)},
lifting state up to \lstinline{Debugger}.
\par

Other than methods to handle state and display components,
\lstinline{Debugger} also contains the Socket.IO client that is used
to communicate with the actual debugger running in the collaborative
debugger's backend.  Socket.IO is a library that extends WebSockets.
It and this communication process will be discussed in detail in
(\ref{rrmessageserver}).  Similarly to how \lstinline{onLogin(name)}
is passed as a property to \lstinline{LoginForm}, the Socket.IO client
is passed to the most important component in the frontend,
\lstinline{RRTerm}, when it is instantiated.

\subsection{The Collaborative Interface to rr}

As shown in Figure \ref{frontend:rrterm}, the primary piece of the
collaborative debugger's frontend is an interface to the rr debugger
backend.  This interface currently consists of a view of the user's
source code and a terminal interface to rr.  These are contained
within the most important class in the frontend: \lstinline{RRTerm}.
This section will outline the process of sending an rr command to the
backend and displaying a response.
\par

In debugging the \lstinline{memoization} example program, users will
want to display the call stack from within each implementation of the
nth fibonacci number function.  To stop in \lstinline{fib_rec}, the
recursive implementation, one user will want to set a breakpoint at
line 52 in the source file.  To do this they will type the command
\lstinline{break 52} and hit enter.  The command will be sent to the
backend, and all users in the same debug session (17734 in Figure
\ref{frontend:rrterm}) will see the debuggers response:\\
\lstinline{Breakpoint 1 at 0x561c8fede303: file memo.c, line 52.}\\
Users who did not send the command will see it displayed along with
the response.

\subsubsection{Sending a Command}

\lstinline{RRTerm}'s state is instantiated as such:

\begin{lstlisting}[language=Javascript,basicstyle=\linespread{0.5}\ttfamily,caption={RRTerm's State},captionpos=b]
  this.state = {command: '',
                  name: '',
                  linum: 0,
                  code: ''};
\end{lstlisting}

\lstinline{name}, \lstinline{linum}, and \lstinline{code} store
information about the current source file being debugged.
\lstinline{command} is the most important field for examing how
commands are sent to the backend, as it holds the current command.
\textit{Every time} the user types a character, \lstinline{command} is
updated to reflect the current state of the Xterm.js component.  The
code to handle keystrokes is shown below:

\begin{lstlisting}[language=Javascript,basicstyle=\linespread{0.5}\ttfamily,caption={Handling Keystrokes},captionpos=b]
  this.term.on('key', (key, ev) => {
    const printable = !ev.altKey && !ev.ctrlKey && !ev.metaKey;
    if (ev.key === 'Enter') {
      // Disable input while we wait for the response from rr
      this.term.disableStdin = true;
      const c = this.state.command;
      this.sendCommand(c);
      this.setState({command: ''});
    } else if (ev.key === 'Backspace') {
      // Do not delete the prompt
      if (this.term.buffers._activeBuffer.x > prompt_length) {
        this.term.write('\b \b');
        this.updateCommand(this.state.command, 'delete');
      }
    } else if (ev.key === 'ArrowUp') {
    } else if (ev.key === 'ArrowDown') {
    } else if (printable) {
      this.term.write(key);
      this.updateCommand(this.state.command, key);
    }
  });
\end{lstlisting}

Whenever the user types a printable key, the key is written to the
terminal and \lstinline{command} is updated using the
\lstinline{updateCommand} method.  If a user types \lstinline{brk},
\lstinline{updateCommand} would be called three times, once for each
keystroke.  Upon realizing they made a mistake when typing
\lstinline{break 52}, the user would likely delete a character.  As
long as the prompt (\lstinline{(rr)} followed by a space) would not be
overwritten, the terminal is updated and \lstinline{updateCommand} is
called again.  When the user hits the enter key the command is finally
sent to the backend using \lstinline{sendCommand}.  To mimick the
behavior of a local rr debug session and to prevent garbled input,
writing to the terminal is disabled until a response is recieved.

\subsubsection{Receiving a Response}

All users connected to the same debug session will recieive responses
to the commands sent by one user.  This is how collaboration is
achieved in the collaborative debugger.  The JSON response object to
\lstinline{break 52} recieved by all users is shown below:

\begin{lstlisting}[basicstyle=\linespread{0.5}\ttfamily,caption={JSON Response},captionpos=b]
{
  channel: ``11734''
  command: ``break 52''
  from: ``1ed644ad87d44cd68121de3daa65adb4''
  response: {
    output: ``Breakpoint 2 at 0x561c8fede303: file memo.c, line 52.''
    source: {
      file_name: null
      current_line: ``87''
      contents: null
    }
  }
}
\end{lstlisting}

The command has been issued from a breakpoint set at \lstinline{main},
which is at line 87 in \lstinline{memo.c}.  Since the source file
hasn't changed, \lstinline{source.file_name} and
\lstinline{source.contents} are set to \lstinline{null}.
\lstinline{output} contains the response from rr, while
\lstinline{channel}, \lstinline{command}, and \lstinline{from} contain
information necessary to display commands correctly for all users.

\begin{lstlisting}[language=Javascript,basicstyle=\linespread{0.5}\ttfamily,caption={Processing a Response},captionpos=b]
  this.props.socket.on('rr_response', (data) => {
    if(data.from !== this.props.socket.id){
      // Display commands from other clients
      this.term.disableStdin = true;
      this.term.clear();
      this.term.write(data.command);
    };
    // Write a newline, but no prompt
    this.term.writeln('');
    // Write each line of the output to the terminal
    data.response.output.split(/\n+/).forEach((l) => {
      this.term.writeln(l);
    });
    this.term.prompt();
    // Re-enable input.  See above and below.
    this.term.disableStdin = false;
    // If an error has occured, there wont be any new
    // information about the location in the source file.
    if(data.response.source != null) {
      this.updateSource(data.response.source);
    }
  });
\end{lstlisting}

When a response is received, \lstinline{RRTerm} first checks to see if
the response has originated from a different user.  If so, input is
disabled, the current line is cleared, and the command that
corresponds to the response is written to the terminal.  Each line of
the response is then written to the terminal.  Finally, input is
renabled, having been disabled either to write commands received from
another user, or when the command was first sent.
\lstinline{updateSource} is called to update the Monaco editor source
file view.  Since \lstinline{break 52} was sent from a location
already inside \lstinline{memo.c}, the source file name and contents
are not updated, though the line number is changed.

\subsection{Considerations} \label{vistools}

The intention of much of the code described above, such as that to
disable input after a command has been sent, is to mimic rr in a way
that is invisible to the user.  The user experience is intended to be
that of using a fully-fledged debugger collaboratively.  The source
code view allows users to easily see what code they are working on
together, and the terminal updates quickly whenever a user enters a
command.
\par

Care has been taken to design the frontend in a modular fashion so
that it can be easily extended in the future.  By sending additional
data about program state in the response and creating new componenets,
visualization tools could be easily added.  If further authentication
features or the ability to edit source code are desired in the future,
relevant components can be updated without the need to change the
entire application.  Since the collaborative debugger depends heavily
on interactions between it's frontend and backend, similar care has
been taken in designing the backend, which will be described in the
following sections.

@@

\subsection{RR Message Server}\label{rrmessageserver}

@@

\subsection{Considerations}

@@

\subsection{Considerations}

The Frontend/API server connects all components of the collaborative
debugger together.  It authenticates users through the database, joins
the frontend and Kubernetes cluster through the dynamic creation of
Pods, and starts the interaction between users and the programs they
wish to debug.  Care has been taken to make it modular and extensible
within reason, and additional authenication or pod creation features
should be easy to add.
\par

The experience provided by invisible components such as the
Frontend/API Server is just as important as that provided by the
frontend webapp.  All components of the collaborative debugger
hopefully come together to provide an experience that facilitates the
teaching and learning of debugging.

@@

\subsection{Features to Add}

The current iteration of the collaborative debugger succesfully
enables collaboration between students.  Significant effort has been
taken to design the debugger in a way that makes collaboration
seemless.  The collaborative debugger hopefully fulfills it's first
requirment---to encourage collaboration.
\par

Though the collaborative debugger makes it easy for teachers to design
and distribute debugging lessons, more work is necessary to fully
fulfill the second goal---that of enhancing students' understanding of
debugging and of programs being debugged.
\par

The most significant addition towards this goal would be that of
frontend vizualization tools.  A simple visualization of the stack and
registers would greatly expand students' understanding of the program
being debugged in all three example sessions currently included with
the collaborative debugger.  gdb/mi supports commands which give
structured output of both these aspects of the execution space, and
the frontend has been designed to make adding visualizations of them
simple.  Adding visualization features will bring the collaborative
debugger much closer towards hopefully fulfilling it's second
requirement.

\subsection{Testing Efficacy}

Though a review of the research seems to indicate that a tool like the
collaborative debugger should help make students more effective at
debugging, further study is needed to draw conclusions.  First, more
research is necessary to determine definitively if teaching debugging
collaboratively is more effective than teaching it individually.
Second, the collaborative debugger needs to be thoroughly tested to
determine whether it is an ideal platform for collaborative debugging.
\par

A study similar to those undertaken in \cite{10.1145/1026487.1008043},
\cite{10.1145/1145287.1145293}, and \cite{10.1145/3361721.3361724}
could be done to test both these aspects.  Students participating in
an introductory systems programming class could be split into three
groups.  Group one could study debugging individually, using rr by
themselves to progress through debugging sessions designed by the
instructor.  Group two could study debugging in pairs, relying on
existing tools such as video-conferencing software to facilitate
collaboration.  In the final group, pairs would use the collaborative
debugger to work together on the same debugging sessions as groups one
and two.  Students' individual and pair performance in debugging both
unseen and self-generated programs could then be tested.

\subsection{Conclusions}

The collaborative debugger aims to enourage collaboration and build
students knowledge of debugging through careful design.  Many
components of a distributed system come together to create a platform
that hopefully benefits both teachers and students.  Through
deliberate design, the collaborative debugger can conceivably be the
ideal platform to teach and learn debugging, both remotely and
in-person.
